% =============================================================================
\section{Experimental Design and Empirical Validation Plan}\label{sec:experiments}
% =============================================================================

We propose a comprehensive experimental program to estimate, compare, and validate ECL across genomic models and tasks. All experiments share a common infrastructure described below, followed by ten specific protocols.

\subsection{Models}\label{sec:exp_models}

We evaluate six models spanning distinct architectural families:
\textbf{Enformer}~\citep{Avsec2021Enformer} (CNN + Transformer, 196~kb),
\textbf{Borzoi}~\citep{Linder2025Borzoi} (CNN + Transformer, 524~kb),
\textbf{HyenaDNA-large}~\citep{Nguyen2023HyenaDNA} (implicit convolution, 1~Mb),
\textbf{Caduceus}~\citep{Schiff2024Caduceus} (BiMamba SSM, 131~kb),
\textbf{Evo~2}~\citep{Brixi2025Evo2} (StripedHyena + Transformer, 1~Mb),
and \textbf{DNABERT-2}~\citep{Zhou2023DNABERT2} (BERT with ALiBi, $\sim$3~kb).

\subsection{Datasets and loci}

We sample 500 loci from each of three classes: (i)~promoter-centered (TSS $\pm$ half-window, from GENCODE), (ii)~enhancer-centered (ENCODE-annotated distal enhancers), (iii)~random intronic/intergenic controls. All loci are drawn from held-out chromosomes (chr8, chr9 for human hg38). Functional genomic tracks from ENCODE~\citep{ENCODE2012}, Roadmap Epigenomics~\citep{Kundaje2015Roadmap}, and GTEx~\citep{GTEx2017} provide ground truth.

\subsection{Perturbation mechanisms}

We compare four perturbation types: (i) random substitution ($\Pi^{\mathrm{sub}}$), (ii) dinucleotide shuffle ($\Pi^{\mathrm{shuf}}$), (iii) 3-mer Markov resampling ($\Pi^{\mathrm{Markov}}$), (iv) generative infilling ($\Pi^{\mathrm{gen}}$, using a small masked LM). ECL is reported for each perturbation type.

\subsection{Experiment 1: Influence profile estimation}

Estimate the binned influence profile $\widehat{\cI}(d;r)$ for each model at each locus class using \cref{alg:binned} with $n=100$ sequences and $m(d)=10$ positions per distance bin. Plot on log-scale.

\paragraph{Proposed Figure~1.}
Six-panel figure: log-scale $\widehat{\cI}(d;r)$ vs.\ distance $d$ (bp) for each model, averaged over promoter-centered loci. One column per perturbation type (showing two: random substitution and dinucleotide shuffle). Shaded bands show $\pm$1 SE. Expected: models with attention show slower decay than pure CNNs; SSM models show intermediate decay.

\paragraph{Proposed Figure~2.}
Same format for enhancer-centered loci. Expected: different profile shape reflecting enhancer-specific regulatory architecture.

\subsection{Experiment 2: ECL estimation and cumulative influence}

Compute cumulative influence curves and ECL estimates at $\beta \in \{0.5, 0.8, 0.9, 0.95, 0.99\}$ for all models.

\paragraph{Proposed Figure~3.}
Cumulative influence fraction $\widehat{\cI}_{\le \ell}(r)/\widehat{\cI}_{\mathrm{tot}}(r)$ vs.\ radius $\ell$ (bp), all models overlaid, with bootstrap 95\% CIs. Horizontal lines mark $\beta$ thresholds. Vertical lines mark corresponding ECL estimates.

\paragraph{Proposed Table~2.}
\begin{table}[H]
\centering
\caption{Proposed format: ECL$_\beta$ estimates (bp) with 95\% bootstrap CIs across models and locus classes. Values to be filled upon experimentation.}
\label{tab:ecl_estimates}
\small
\begin{tabular}{@{}ll ccccc @{}}
\toprule
\textbf{Model} & \textbf{Locus class} & $\beta{=}0.5$ & $\beta{=}0.8$ & $\beta{=}0.9$ & $\beta{=}0.95$ & $\beta{=}0.99$ \\
\midrule
Enformer & Promoter & --- & --- & --- & --- & --- \\
Enformer & Enhancer & --- & --- & --- & --- & --- \\
Enformer & Intronic & --- & --- & --- & --- & --- \\
\midrule
Borzoi & Promoter & --- & --- & --- & --- & --- \\
Borzoi & Enhancer & --- & --- & --- & --- & --- \\
Borzoi & Intronic & --- & --- & --- & --- & --- \\
\midrule
HyenaDNA & Promoter & --- & --- & --- & --- & --- \\
Caduceus & Promoter & --- & --- & --- & --- & --- \\
Evo~2 & Promoter & --- & --- & --- & --- & --- \\
DNABERT-2 & Promoter & --- & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Proposed Table~3.}
\begin{table}[H]
\centering
\caption{Proposed format: context utilization ratio $\ECL_{0.9}$/Nominal for each model.}
\label{tab:utilization}
\small
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Model} & \textbf{Nominal (bp)} & $\widehat{\ECL}_{0.9}$ \textbf{(bp)} & \textbf{Utilization (\%)} \\
\midrule
Enformer & 196,608 & --- & --- \\
Borzoi & 524,288 & --- & --- \\
HyenaDNA & 1,000,000 & --- & --- \\
Caduceus & 131,072 & --- & --- \\
Evo~2 & 1,000,000 & --- & --- \\
DNABERT-2 & $\sim$3,000 & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Experiment 3: Directional ECL}

Compute upstream and downstream influence profiles and $\ECL_{0.9}^-$, $\ECL_{0.9}^+$ at promoter-centered loci for Enformer and Borzoi.

\paragraph{Proposed Figure~4.}
Two-panel figure: upstream (negative distances) and downstream (positive distances) influence profiles for Enformer and Borzoi. Expected: asymmetry reflecting promoter-upstream enhancer effects.

\subsection{Experiment 4: Perturbation sensitivity}

Compare ECL estimates across perturbation types for each model.

\paragraph{Proposed Table~4.}
\begin{table}[H]
\centering
\caption{Proposed format: ECL$_{0.9}$ (bp) under different perturbation types.}
\label{tab:perturbation_sensitivity}
\small
\begin{tabular}{@{}l cccc @{}}
\toprule
\textbf{Model} & $\Pi^{\mathrm{sub}}$ & $\Pi^{\mathrm{shuf}}$ & $\Pi^{\mathrm{Markov}}$ & $\Pi^{\mathrm{gen}}$ \\
\midrule
Enformer & --- & --- & --- & --- \\
Borzoi & --- & --- & --- & --- \\
HyenaDNA & --- & --- & --- & --- \\
Caduceus & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Proposed Figure~5.}
Scatter plot of ECL$_{0.9}$(shuffle) vs.\ ECL$_{0.9}$(substitution) across 500 promoter loci, one point per locus, colored by model. Expected: positive correlation with systematic offset.

\subsection{Experiment 5: Trained vs.\ random weights}

Compare influence profiles for Enformer with trained weights vs.\ randomly initialized weights (same architecture).

\paragraph{Proposed Figure~6.}
Overlay of influence profiles for trained vs.\ random Enformer. Expected: random weights produce a profile determined solely by architecture (TRF), while trained weights show data-driven ECL shaped by learned regulatory features.

\subsection{Experiment 6: Multi-scale block ECL}

Compute ECL at block sizes $b \in \{1, 32, 128, 512\}$ bp for Borzoi.

\paragraph{Proposed Table~5.}
\begin{table}[H]
\centering
\caption{Proposed format: multi-scale block ECL for Borzoi at promoter loci.}
\label{tab:multiscale}
\small
\begin{tabular}{@{}l cccc @{}}
\toprule
$\beta$ & $b=1$ bp & $b=32$ bp & $b=128$ bp & $b=512$ bp \\
\midrule
0.5 & --- & --- & --- & --- \\
0.9 & --- & --- & --- & --- \\
0.95 & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Experiment 7: Locus-class heterogeneity}

Compute ECL$_{0.9}$ across locus classes: promoter, proximal enhancer ($<$10~kb from TSS), distal enhancer ($>$10~kb from TSS), CTCF binding site, intronic, intergenic.

\paragraph{Proposed Figure~7.}
Violin plots of $\widehat{\ECL}_{0.9}$ distribution across locus classes for Enformer and Borzoi. Expected: distal enhancer loci have larger ECL than promoter loci.

\paragraph{Proposed Figure~8.}
Scatter plot of ECL$_{0.9}$ vs.\ gene length (bp), colored by expression level. Expected: longer genes may have larger ECL if models capture gene-body-level regulation.

\subsection{Experiment 8: Model comparison}

Paired comparison of ECL across models at matched loci using the permutation test (\cref{alg:compare}).

\paragraph{Proposed Figure~9.}
Paired difference plot (Borzoi $-$ Enformer) of ECL$_{0.9}$ at 500 matched loci, with density and mean $\pm$ CI. Stratified by locus class.

\paragraph{Proposed Table~6.}
\begin{table}[H]
\centering
\caption{Proposed format: pairwise model comparison. Each cell: mean $\Delta$ECL$_{0.9}$ (row $-$ column) with permutation test $p$-value.}
\label{tab:pairwise}
\small
\begin{tabular}{@{}l cccc @{}}
\toprule
& Enformer & Borzoi & HyenaDNA & Caduceus \\
\midrule
Enformer & --- & --- & --- & --- \\
Borzoi & --- & --- & --- & --- \\
HyenaDNA & --- & --- & --- & --- \\
Caduceus & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Experiment 9: Biological validation}

For known long-range enhancer--gene pairs, check whether model ECL reaches the known regulatory distance.

\paragraph{Proposed Figure~10.}
For the SHH/ZRS system ($\sim$1~Mb), the MYC super-enhancer ($\sim$1.7~Mb), and the SOX9 regulatory desert ($\sim$1~Mb), plot the full influence profile centered on the target gene's TSS and mark the known enhancer position. Overlay for Enformer, Borzoi, and Evo~2.

\paragraph{Proposed Table~7.}
\begin{table}[H]
\centering
\caption{Proposed format: biological validation against known long-range interactions.}
\label{tab:validation}
\small
\begin{tabular}{@{}l r cccc @{}}
\toprule
\textbf{Locus} & \textbf{Known dist.\ (kb)} & \textbf{Enformer} & \textbf{Borzoi} & \textbf{Evo~2} & \textbf{Detected?} \\
\midrule
SHH/ZRS & $\sim$1,000 & --- & --- & --- & --- \\
MYC enhancer & $\sim$1,700 & --- & --- & --- & --- \\
SOX9 desert & $\sim$1,000 & --- & --- & --- & --- \\
HBB/LCR & $\sim$60 & --- & --- & --- & --- \\
PAX6/ELP4 & $\sim$150 & --- & --- & --- & --- \\
SHH/MACS1 & $\sim$900 & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Experiment 10: Interaction ECL and gNIAH}

\paragraph{Interaction ECL.}
Compute pairwise interaction influence (\cref{eq:interaction}) at the HBB locus control region, a well-characterized system with cooperative enhancer elements spanning $\sim$60~kb.

\paragraph{Proposed Figure~11.}
Heatmap of $\widehat{\cI}_{\mathrm{int}}(i,j;r)$ at the HBB locus, showing synergistic blocks between the locus control region elements and the $\beta$-globin gene.

\paragraph{gNIAH protocol.}
Insert CTCF, GATA, and SP1 motifs at distances $d \in \{1, 2, 5, 10, 20, 50, 100, 200\}$ kb from center in dinucleotide-shuffled backgrounds. Measure gNIAH sensitivity (\cref{eq:gniah}) for each model.

\paragraph{Proposed Figure~12.}
gNIAH sensitivity vs.\ distance for three motifs $\times$ six models. Expected: motif-specific decay rates; CTCF detected at longer range than SP1.
